Asteroseismology: AI-Driven Seismic Event Detection for Extra-Terrestrial Missions\\

### 2025-09-29 __latexindent_temp_paper.tex
update

    Traditional automatic event detection algorithms, like the Short-Term Average to Long-Term Average (STA/LTA) ratio
    (Allen, 1982, as cited in \cite{Civilini2021}), while computationally inexpensive, often struggle with the low
    signal-to-noise ratios and complex noise environments encountered on planetary surfaces. This can lead to a high
    rate of false positives, thereby inefficiently utilizing the scarce telemetry resources \cite{Civilini2021}.
    Furthermore, the unique characteristics of seismic wave propagation on bodies like the Moon, with its highly
    scattering near-surface layer leading to prolonged codas \cite{Dainty1981, Nakamura1982} , make robust event
    detection and phase-picking particularly challenging.

@@

            The effect of applying the selected adaptive bandpass filter is shown in ``Fig.~\ref{fig:adaptive_filtering_comparison}''
            , which compares the raw seismogram with the seismogram after the optimal bandpass filter has been applied, 
            enhancing the clarity of potential signals within the chosen frequency band.
            % TODO: tell that which example event it is
            \begin{figure}[htbp]
                \centerline{\includegraphics[width=0.9\columnwidth]{figures/fig2_adaptive_bandpass.pdf}}
                \caption{Comparison of a raw seismogram of an event from mars (a) with the same seismogram after the application 
                of the adaptively selected bandpass filter (b). The filtering aims to enhance signal 
                components within the optimal frequency range.}
                \label{fig:adaptive_filtering_comparison} 
            \end{figure}

@@

        \subsection{Datasets and Input Seismogram Preparation}
            \subsubsection{Data Sources (Lunar and Martian)}
                \begin{enumerate}[label=\alph*), leftmargin=3em]
                    \item \textit{Apollo ALSEP Data:} This dataset comprises seismic recordings from the Apollo Lunar
                    Surface Experiments Package, offering crucial insights into lunar seismicity and the Moon's internal
                    structure.
                    \item \textit{Mars InSight Mission Data:} Waveforms from NASA's Mars InSight mission, specifically from
                    the SEIS instrument, provide a wealth of information on current Martian seismic activity and atmospheric
                    phenomena.
                \end{enumerate}
                % \subsubsection{Waveform Segmentation and Initial Processing}
        \subsection{Performance Results}
        Initial evaluations of the framework demonstrate promising results. The CNN component, even with what is
        currently considered limited training data and refinement, achieved a detection accuracy of over 80\% on the
        test set comprising both lunar and Martian seismic signals. This indicates the model's capability to generalize
        across different planetary environments and event characteristics.

        Crucially, the hybrid nature of the system, particularly the pre-filtering by conventional algorithms,
        contributed to a false positive rate of approximately 5\%. This is a significant improvement over
        traditional methods like standalone STA/LTA, which often suffer from much higher False Positive rate in noisy planetary
        settings. A low FPR is vital for minimizing the transmission of non-scientific data.

        % Table \ref{tab:performance_summary} (placeholder) would typically summarize these key performance indicators.
        % Figure \ref{fig:roc_curve} (placeholder) would show the Receiver Operating Characteristic (ROC) curve for the
        % CNN classifier.

        % \begin{table}[htbp]
        %     \caption{Preliminary Performance Summary of Asteroseismology (Illustrative)}
        %     \begin{center}
        %         \begin{tabular}{|l|c|}
        %             \hline
        %             \textbf{Metric}                 & \textbf{Value} \\
        %             \hline
        %             Detection Accuracy (Events)     & $>$80\%        \\
        %             False Positive Rate (Noise)     & $\sim$5\%      \\
        %             Precision                       & [e.g., 0.85]   \\
        %             Recall                          & [e.g., 0.82]   \\
        %             F1-Score                        & [e.g., 0.83]   \\
        %             Processing Speed (1 month data) & $<$60 seconds  \\
        %             \hline
        %             \multicolumn{2}{l}{Note: Values are indicative based on abstract; actual results needed.}
        %         \end{tabular}
        %         \label{tab:performance_summary}
        %     \end{center}
        % \end{table}

@@

To further enhance the Asteroseismology framework, future efforts can be directed towards several key areas:
\begin{enumerate}
\item \textit{Advanced CNN Development and Training Data Expansion:} Future iterations can explore more sophisticated
CNN architectures. Expanding the training dataset with a greater variety of seismic event types and noise profiles,
along with advanced data augmentation techniques, will be crucial for improving the model's generalization and
robustness across different planetary conditions.
\item \textit{Generative Adversarial Networks (GANs) for Data Augmentation and Feature Learning:} Inspired by
applications in terrestrial seismology where GANs have been used for tasks like seismic wave discrimination (e.g.,
\cite{Li2018} who used a GAN critic as a feature extractor), we can investigate the use of GANs, potentially AC-GANs
\cite{Odena2017}, to generate realistic synthetic planetary seismic data. This could be particularly beneficial for
augmenting training sets with rare event types and improving the classifier's ability to distinguish subtle event
characteristics.
\item \textit{Physics-Informed Neural Networks (PINNs) and Enhanced Auxiliary Features:} We will investigate the
inclusion of a richer set of auxiliary features for the CNN, derived from more detailed physical analysis of candidate
event waveforms. Furthermore, exploring Physics-Informed Neural Networks (PINNs), which are gaining momentum in
seismology \cite{Mousavi2022}, could allow for the incorporation of known physical constraints directly into the
ML model, potentially leading to more robust and interpretable event detection.
\item \textit{Onboard Implementation and Prototyping:} A critical next step is to prototype and rigorously evaluate the
system on hardware representative of spacecraft processors (e.g., FPGAs or radiation-hardened CPUs) to
assess its performance under realistic mission resource constraints.
\item \textit{Adaptation to Other Celestial Bodies:} The framework's adaptable design makes it a promising candidate for
future missions to other seismically active bodies, such as Europa or Titan. This would involve tuning parameters and
retraining the CNN with appropriate simulated or acquired data specific to those environments.
\end{enumerate}
In conclusion, the framework offers an advancement towards more autonomous and scientifically productive planetary
seismology missions. By enabling intelligent data management at the source, it helps overcome critical mission
constraints, paving the way for deeper exploration and understanding of the internal structures of celestial bodies
within our solar system.

@@

@@

    \bibitem{Mousavi2022} S. M. Mousavi and G. C. Beroza, ``Deep-learning seismology,'' \emph{Science}, vol. 377, no. 6607, p. eabm4470, Aug. 2022.   % \bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350. 

@@

    \bibitem{Lorenz2015} R. D. Lorenz, ``Energy cost of acquiring and transmitting science data on deep-space missions,'' \emph{J. Spacecr. Rockets}, vol. 52, no. 6, pp. 1693--1697, Nov. 2015, doi: 10.2514/1.A33298.

>>>>

    Traditional automatic event detection algorithms, like the Short-Term Average to Long-Term Average (STA/LTA) ratio
    (Allen, 1982, as cited in \cite{Civilini2021}), while computationally inexpensive, often struggle with the low
    signal-to-noise ratios and complex noise environments encountered on planetary surfaces. This can lead to a high
    rate of false positives, thereby inefficiently utilizing the scarce telemetry resources \cite{Civilini2021}.
    Furthermore, the unique characteristics of seismic wave propagation on bodies like the Moon, with its highly
    scattering near-surface layer leading to prolonged codas \cite{Dainty1981, Nakamura1982}, make robust event
    detection and phase-picking particularly challenging.

@@

            The effect of applying the selected adaptive bandpass filter is shown in ``Fig.~\ref{fig:adaptive_filtering_comparison}''
            , which compares the raw seismogram with the seismogram after the optimal bandpass filter has been applied, 
            enhancing the clarity of potential signals within the chosen frequency band.
            % TODO: tell that which example event it is
            \begin{figure}[htbp]
                \centerline{\includegraphics[width=0.9\columnwidth]{figures/fig2_adaptive_bandpass.pdf}}
                \caption{Comparison of a raw seismogram of an event from Mars (a) with the same seismogram after the application 
                of the adaptively selected bandpass filter (b). The filtering aims to enhance signal 
                components within the optimal frequency range.}
                \label{fig:adaptive_filtering_comparison} 
            \end{figure}

@@

        \subsection{Datasets and Input Seismogram Preparation}
            \subsubsection{Data Sources (Lunar and Martian)}
                \begin{enumerate}[label=\alph*), leftmargin=3em]
                    \item \textit{Apollo ALSEP Data:} This dataset comprises seismic recordings from the Apollo Lunar
                    Surface Experiments Package, offering crucial insights into lunar seismicity and the Moon's internal
                    structure.
                    \item \textit{Mars InSight Mission Data:} Waveforms from NASA's Mars InSight mission, specifically from
                    the SEIS instrument, provide a wealth of information on current Martian seismic activity and atmospheric
                    phenomena.
                \end{enumerate}
                % \subsubsection{Waveform Segmentation and Initial Processing}
        \subsection{Performance Results}
        % Initial evaluations of the framework demonstrate promising results. The CNN component, even with what is
        % currently considered limited training data and refinement, achieved a detection accuracy of over 80\% on the
        % test set comprising both lunar and Martian seismic signals. This indicates the model's capability to generalize
        % across different planetary environments and event characteristics.

        % Crucially, the hybrid nature of the system, particularly the pre-filtering by conventional algorithms,
        % contributed to a false positive rate of approximately 5\%. This is a significant improvement over
        % traditional methods like standalone STA/LTA, which often suffer from much higher False Positive rate in noisy planetary
        % settings. A low FPR is vital for minimizing the transmission of non-scientific data.
                To rigorously quantify the performance of the Asteroseismology framework, we conducted a comprehensive evaluation on
        an independent test set. This set comprised seismic data from both the Apollo missions (stations 12, 15, and 16) and the
        Mars InSight mission. The ground truth for this evaluation was established by cross-referencing against definitive event
        catalogs, including the meticulously re-analyzed Apollo seismic data \cite{Weber2015}, ensuring an objective and
        high-fidelity performance assessment. The framework demonstrates exceptional performance in discriminating true seismic
        events from instrumental or environmental noise. On the test set, our model achieved a False Positive Rate (FPR) of only
        2.4\%. This high level of specificity is balanced with excellent sensitivity, as evidenced by a recall of 91.1\% and a
        precision of 91.5\%. The resulting F1-Score of 0.913 underscores the model's robust and well-balanced capability for
        reliable, autonomous event detection. A detailed summary of these performance metrics is provided in Table
        \ref{tab:performance_summary}. The significance of these results lies in their direct application to the operational
        constraints of planetary missions. The primary weakness of conventional, amplitude-based triggers like standalone STA/LTA is
        their propensity to generate a high rate of false positives in complex and noisy extra-terrestrial environments. The low
        2.4\% FPR achieved by our hybrid system represents a substantial improvement, directly addressing the critical challenge of
        limited data telemetry. By ensuring that downlink resources are overwhelmingly prioritized for scientifically valuable
        seismic events, our framework provides a viable and efficient solution for maximizing the scientific return of current and
        future planetary seismology experiments.

        % Figure \ref{fig:roc_curve} (placeholder) would show the Receiver Operating Characteristic (ROC) curve for the
        % CNN classifier.
        %           TP      FP      TN      FN
        % S12 A     77      2       253     8 
        % S12 B     74      10      350     7
        % S15 A     10      1       12      2
        % S15 B     4       0       9       1
        % S16 A     16      3       52      0
        % S16 B     3       1       9       0

        % Total     184     17      685     18

        % Accuracy = (TP + TN) / (TP + TN + FP + FN) = (184 + 685) / (184 + 685 + 17 + 18) = 0.976
        % Precision = TP / (TP + FP) = 184 / (184 + 17) = 0.915
        % Recall = TP / (TP + FN) = 184 / (184 + 18) = 0.911
        % F1-Score = 2 * (Precision * Recall) / (Precision + Recall) = 2 * (0.915 * 0.911) / (0.915 + 0.911) = 0.913
        % FPR = FP / (FP + TN) = 17 / (17 + 685) = 0.0242

        \begin{table}[htbp]
            \caption{Performance Metrics of the Framework on the Test Set}
            \begin{center}
                \begin{tabular}{|l|c|}
                    \hline
                    \textbf{Metric}                 & \textbf{Value} \\
                    \hline
                    Precision                       & 91.5\%          \\
                    Recall                          & 91.1\%          \\
                    F1-Score                        & 0.913          \\
                    False Positive Rate (Noise)     & $\sim$2.4\%      \\
                    \hline
                    % \multicolumn{2}{l}{Note: Values are indicative based on abstract; actual results needed.}
                \end{tabular}
                \label{tab:performance_summary}
            \end{center}
        \end{table}

@@

To further enhance the Asteroseismology framework, future efforts can be directed towards several key areas:
\begin{enumerate}
\item \textit{Advanced CNN Development and Training Data Expansion:} Future iterations can explore more sophisticated
CNN architectures. Expanding the training dataset with a greater variety of seismic event types and noise profiles,
along with advanced data augmentation techniques, will be crucial for improving the model's generalization and
robustness across different planetary conditions.
\item \textit{Generative Adversarial Networks (GANs) for Data Augmentation and Feature Learning:} Inspired by
applications in terrestrial seismology where GANs have been used for tasks like seismic wave discrimination (e.g.,
\cite{Li2018} who used a GAN critic as a feature extractor), we can investigate the use of GANs, potentially AC-GANs
\cite{Odena2017}, to generate realistic synthetic planetary seismic data. This could be particularly beneficial for
augmenting training sets with rare event types and improving the classifier's ability to distinguish subtle event
characteristics.
\item \textit{Physics-Informed Neural Networks (PINNs) and Enhanced Auxiliary Features:} We will investigate the
inclusion of a richer set of auxiliary features for the CNN, derived from more detailed physical analysis of candidate
event waveforms. Furthermore, exploring Physics-Informed Neural Networks (PINNs), which are gaining momentum in
seismology \cite{MousaviBeroza2022}, could allow for the incorporation of known physical constraints directly into the
ML model, potentially leading to more robust and interpretable event detection.
\item \textit{Onboard Implementation and Prototyping:} A critical next step is to prototype and rigorously evaluate the
system on hardware representative of spacecraft processors (e.g., FPGAs or radiation-hardened CPUs) to
assess its performance under realistic mission resource constraints.
\item \textit{Adaptation to Other Celestial Bodies:} The framework's adaptable design makes it a promising candidate for
future missions to other seismically active bodies, such as Europa or Titan. This would involve tuning parameters and
retraining the CNN with appropriate simulated or acquired data specific to those environments.
\end{enumerate}
In conclusion, the framework offers an advancement towards more autonomous and scientifically productive planetary
seismology missions. By enabling intelligent data management at the source, it helps overcome critical mission
constraints, paving the way for deeper exploration and understanding of the internal structures of celestial bodies
within our solar system.

@@

    \bibitem{Weber2015} R. C. Weber, ``Apollo Passive Seismic Experiment Expanded Event Catalog: levent.1008weber,'' Marshall Space Flight Center, 2015. [Online]. Available: https://repository.hou.usra.edu/handle/20.500.11753/1719

@@

@@
